---
title: "Cancer Evenness Replication"
author: "Benjamin Cocanougher"
date: "August 17, 2015"
output: html_document
---

The data used for this analysis comes from the [International Agency for Research on Cancer](http://ci5.iarc.fr/CI5I-X/Pages/download.aspx). Specifically we will look at the yellow volume VII data. The CI5VII.csv file can be downloaded [here](http://ci5.iarc.fr/CI5I-X/old/vol7/CI5-VII_software.zip) from the IARC. Alternatively, the files, including this analysis, are available in this github repo.


```{r, echo = TRUE, cache=TRUE}

# You will need to edit this to the working 
# directory containing the CI5VII.csv file
setwd("/Users/bencocanougher/Desktop/CI5-VIId")

if (!exists("cancer")) {
  cancer <- read.csv("CI5VII.csv")
}

colnames(cancer) <- c("pop", "sex", "cancer", "ageGroup",
                      "cancerByAge", "PersonYearsatRisk")

str(cancer)
```

GetEvenness function is defined:
```{r}
GetEvenness <- function(vector) {
  n <- sum(vector)
  s <- length(vector)
  print("", quote = F)
  print(c("sum:",n), quote = F)
  print(c("s:",s), quote = F)
  # Basharin's small-sample correction:
  ens <- (s-1)/(2*log(2)*n);
  H <- -sum((vector/n)*log2(vector/n)) + ens
  Hmax <- log2(s) 
  H/Hmax
  #print(c("evenness:", round(H/Hmax,3)), quote = F)
}
```

Next I define a function to subset the data and calculate the evenness.

```{r}
CancerEvenness <- function(ageGroupID) {
  # First subset the cancer data frame by the ageGroup column
  x <- cancer[cancer$ageGroup == ageGroupID, ]
  
  # Then sum all of the data by cancer type
  x2 <- aggregate(. ~ cancer, data = x, FUN=sum)
  
  # Remove cancer types with no data (non-existant 
  # species in population of interest)
  x3 <- x2$cancerByAge[which(x2$cancerByAge != 0)]
  
  # Calculate the evenness and print to the console
  x4 <- GetEvenness(x3)
  print(paste0("Evenness of Age Group ", ageGroupID," = ", x4), quote = FALSE)
}
```


With the above function, we can now easily calculate the evenness for each age group.
```{r}
x <- lapply(1:19, CancerEvenness)
```
